Halving the cost of quantum addition

### 2017-08-28 paper.tex
Applying Austin's comments: - Shortened title to be punchy - Fixed logical-and -> logical-AND - T-cost -> T-count

\title{T gates, temporary ANDs, and halving the cost of quantum addition}
\author{Craig Gidney}
\affiliation{Google, Santa Barbara, CA 93117, USA}
\email{craiggidney@google.com}

@@

\begin{abstract}
We improve the number of T gates needed to perform an $n$-bit adder from $8n + O(1)$ \cite{Cuccaro2004, Amy2013, AustinDiscussionsAndEmails2017} to $4n + O(1)$.
We do so via a ``temporary AND gate" construction, which uses four T gates to store the logical-and of two qubits into an ancilla and zero T gates to later erase the ancilla.

Temporary AND gates are a generally useful tool when optimizing T-cost.
They can be applied to integer arithmetic, modular arithmetic, rotation synthesis, the quantum Fourier transform, Shor's algorithm, Grover oracles, and many other circuits.
Because T gates dominate the cost of quantum computation based on the surface code, and the temporary AND gate is widely applicable, our constructions represent a significant reduction in projected costs of quantum computation.

In addition to our $n$-bit adder circuit with T-cost of $4n + O(1)$, we show how to construct an $n$-bit controlled adder circuit with T-cost of $8n + O(1)$, and a temporary adder that can be computed for the same cost as the normal adder but whose result can be kept until later uncomputed at no T-cost.
\end{abstract}

@@

Our paper is divided into six sections.
In \autoref{sec:introduction}, we motivate the problem and introduce terminology.
\autoref{sec:review} discusses how previous work on optimizing T gate counts improved the cost of matched pairs of Toffolis from 14 to 8.
In \autoref{sec:invest}, we show how to improve the T-cost from 8 to 6 by investing and later recovering an $|A\rangle$ state instead of consuming it to perform T gates.
\autoref{sec:temporary-and} expands on those ideas with an ancilla erasing trick that improves the T-cost from 6 to 4.
This leads to \autoref{sec:circuit-constructions}, where we demonstrate how to use our constructions to create low T-cost circuits for several addition tasks and discuss other possible applications of our constructions.
Finally, \autoref{sec:conclusion} summarizes our contributions, and proves that at least two T gates are needed to perform a temporary AND.

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ctrl{1}&\ctrl{1}&\ctrl{1}&\ctrl{1}&\qw && &&&\qw     &\qw     &\qw     &\ctrl{1}&\qw     &\qw     &\qw     &\qw &\\
      &\ctrl{1}&\ctrl{2}&\ctrl{3}&\ctrl{4}&\qw &&=&&&\qw     &\qw     &\qw     &\ctrl{1}&\qw     &\qw     &\qw     &\qw &\\
      &\targ   &\qw     &\qw     &\qw     &\qw && &&&\ctrl{3}&\ctrl{2}&\ctrl{1}&\targ   &\ctrl{1}&\ctrl{2}&\ctrl{3}&\qw &\\
      &\qw     &\targ   &\qw     &\qw     &\qw && &&&\qw     &\qw     &\targ   &\qw     &\targ   &\qw     &\qw     &\qw &\\
      &\qw     &\qw     &\targ   &\qw     &\qw && &&&\qw     &\targ   &\qw     &\qw     &\qw     &\targ   &\qw     &\qw &\\
      &\qw     &\qw     &\qw     &\targ   &\qw && &&&\targ   &\qw     &\qw     &\qw     &\qw     &\qw     &\targ   &\qw &\\
    }
  }
  \caption{
	The T-cost of $N$ adjacent Toffolis sharing the same controls is $0 \cdot N + O(1)$.
	The marginal T-cost is 0 because each additional Toffoli can be replaced with CNOTs framing a root Toffoli.
  }
  \label{fig:shared-controls}
\end{figure}

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ctrl{1} &\qw & &       & & &\ctrl{1}     &\ctrl{1} &\qw & &   & & &\qw &\qw &\qw      &\qw  &\ctrl{2} &\qw &\qw      &\qw  &\qw &\qw \\
      &\ctrl{1} &\qw & & \cong & & &\qS          &\ctrl{1} &\qw & & = & & &\qw &\qw &\ctrl{1} &\qw  &\qw      &\qw &\ctrl{1} &\qw  &\qw &\qw \\
      &\targ    &\qw & &       & & &\gate{Z}\qwx &\targ    &\qw & &   & & &\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\qw \\
    }
  }
  \caption{
	Starting with \autoref{fig:textbook-toffoli} then dropping T gates on the controls produces an operation with a T-cost of 4 that performs the correct permutation.
	However, the operation introduces phase errors.
  }
  \label{fig:bad-phase-toffoli}
\end{figure}

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw && &&&\ctrl{1}     &\ctrl{1}  &\qw &\qw &\qw    &\qw &\qw     &\ctrl{1}&\ctrl{1}     &\qw & \\
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw &&=&&&\qS          &\ctrl{1}  &\qw &\qw &\qw    &\qw &\qw     &\ctrl{1}&\qSi         &\qw & \\
      &\targ   &\qw &    &\ldots &    &    &\targ   &\qw && &&&\gate{Z}\qwx &\targ     &\qw &    &\ldots &    &        &\targ   &\gate{Z}\qwx &\qw & \\
    }
  }
  \caption{
	When two Toffolis form a compute/uncompute pair, they can cancel each others' phase errors.
	Fixes the problem with the construction in \autoref{fig:bad-phase-toffoli}, and achieves a per-Toffoli T-cost of 4 for paired Toffolis \cite{Barenco1995}.
  }
  \label{fig:cancelled-bad-phase-toffoli}
\end{figure}

A typical $n$-bit quantum adder will contain $2n + O(1)$ Toffoli gates, implying a naive T-cost of $14n + O(1)$.
However, almost all of the Toffoli gates in the first half of an adder are uncomputed by Toffoli gates in the second half.
This allows the T gates on the controls of the Toffoli gates to be omitted, reducing their T-cost from 7 to 4 and the T-cost of addition to $8n + O(1)$.

Even if a Toffoli isn't paired with a second Toffoli that uncomputes its effects, it's still possible to perform the Toffoli with T-cost of by using an ancilla, a measurement, and a conditional fixup operation \cite{Jones2013} (see \autoref{fig:jones-toffoli}).
Our key improvement over \cite{Jones2013} is to notice that the ancilla's lifetime can be extended, so it can be used both when computing and uncomputing a Toffoli.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ctrl{1} &\qw & &   & & &     &\qw &\qw &\qw      &\qw  &\ctrl{3} &\qw &\qw      &\qw  &\qw &\qw  &\qw       &\qw &\qw    &\ctrl{1}       &\qw \\
      &\ctrl{1} &\qw & & = & & &     &\qw &\qw &\ctrl{2} &\qw  &\qw      &\qw &\ctrl{2} &\qw  &\qw &\qw  &\qw       &\qw &\qw    &\gate{Z}       &\qw \\
      &\targ    &\qw & &   & & &     &\qw &\qw &\qw      &\qw  &\qw      &\qw &\qw      &\qw  &\qw &\qw  &\targ     &\qw &\qw    &\qw\cwx        &\qw \\
      &         &    & &   & & & \qO &\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\qSi &\ctrl{-1} &\qH &\meter &\cw\cwx\bullet & \\
    }
  }
  \caption{
	Toffoli construction with T-cost of 4 from \cite{Jones2013}.
  }
  \label{fig:jones-toffoli}
\end{figure}

\section{Warm up: Investing $|A\rangle$ states}
\label{sec:invest}

This paper started with an idea that improved the T-cost of a single Toffoli gate from 7 to 6 in a surprising way, by performing the Toffoli indirectly instead of directly.
We started with a clean ancilla in the $|0\rangle$ state, applied a Toffoli to store the logical-and of the two controls in the ancilla, used the ancilla to control a CNOT onto the intended target, then uncomputed the ancilla.

Initially, this indirect-Toffoli construction appears to have a T-cost of 8.
However, as shown in \autoref{fig:indirect-toffoli}, the last T gate in the circuit is unnecessary.
In fact, it is actively harmful.
Dropping this T gate and the following Hadamard not only reduces the T-cost from 8 to 7, it leaves the ancilla in an $|A\rangle$ state that can be consumed to perform a T gate elsewhere.
This improves the net T-cost to 6.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ctrl{1} &\qw && &&&    &\qw &\qw &\qw      &\qw  &\ctrl{3} &\qw &\qw      &\qw  &\qw &\qw       &\qw &\qw &\qw      &\qw  &\ctrl{3} &\qw &\qw      &\qw  &\qw &\qw &&\\
      &\ctrl{1} &\qw &&=&&&    &\qw &\qw &\ctrl{2} &\qw  &\qw      &\qw &\ctrl{2} &\qw  &\qw &\qw       &\qw &\qw &\ctrl{2} &\qw  &\qw      &\qw &\ctrl{2} &\qw  &\qw &\qw &&\\
      &\targ    &\qw && &&&    &\qw &\qw &\qw      &\qw  &\qw      &\qw &\qw      &\qw  &\qw &\targ     &\qw &\qw &\qw      &\qw  &\qw      &\qw &\qw      &\qw  &\qw &\qw &&\\
      &         &    && &&&\qO &\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\ctrl{-1} &\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\qw 
          \gategroup{4}{6}{4}{10}{.7em}{--} \gategroup{4}{26}{4}{30}{.7em}{--} &|0\rangle & \\
      &         &    && &&&    &    |A\rangle &  & &     &         &    &         &     &    &          &    &    &         &    &         &     &         &     & |A\rangle
    }
  }
  \caption{
	Performing a Toffoli indirectly, by applying a pair of Toffolis to an ancilla and using the intermediate value to control a CNOT onto the actual target, appears to have a T-cost of 8.
	However, the initial T gate is being used to compute an $|A\rangle$ state and the final T gate is being used to uncompute the $|A\rangle$ state.
    Instead of spending $|A\rangle$ states to perform T gates to make and unmake $|A\rangle$ states, we pass in and later recover a single $|A\rangle$ state.
	This ``investment" reduces the net T-cost to 6.
  }
  \label{fig:indirect-toffoli}
\end{figure}

A Toffoli with a T-cost of 6 isn't optimal, but the reduction from 7 was done in an unexpected way.
While searching for other circuits where this optimization might apply, we realized it would be useful when an initial Toffoli was later uncomputed by a second Toffoli.
Instead of computing and uncomputing the ancilla for the first Toffoli, then recomputing and re-uncomputing the ancilla for the second Toffoli, we simply kept the ancilla around until it was time to uncompute the Toffoli.
This halved the T-cost of the pair from 12 to 6.
See \autoref{fig:ancilla-temporary-toffoli}.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw && &&&    &\qw     &\qw  &\ctrl{3}&\qw &\qw     &\qw  &\qw &\qw      &\qw &\qw &\qw    &\qw &\qw &\qw      &\qw &\qw &\qw     &\qw  &\ctrl{3}&\qw &\qw     &\qw &&\\
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw &&=&&&    &\ctrl{2}&\qw  &\qw     &\qw &\ctrl{2}&\qw  &\qw &\qw      &\qw &\qw &\qw    &\qw &\qw &\qw      &\qw &\qw &\ctrl{2}&\qw  &\qw     &\qw &\ctrl{2}&\qw &&\\
      &\targ   &\qw &    &\ldots &    &    &\targ   &\qw && &&&    &\qw     &\qw  &\qw     &\qw &\qw     &\qw  &\qw &\targ    &\qw &    &\ldots &    &    &\targ    &\qw &\qw &\qw     &\qw  &\qw     &\qw &\qw     &\qw &&\\
      &        &    &    &       &    &    &        &    && &&&\qA &\targ   &\qTi &\targ   &\qT &\targ   &\qTi &\qH &\ctrl{-1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{-1}&\qH &\qT &\targ   &\qTi &\targ   &\qT &\targ   &\qw &|A\rangle &\\
    }
  }
  \caption{
	Computing and uncomputing a Toffoli gate with a net T-cost of 6.
	Invests an $|A\rangle$ state when computing the Toffoli, holds an ancilla qubit storing the logical-and of the controls until it's time to uncompute the Toffoli, then recovers the $|A\rangle$ state.
  }
  \label{fig:ancilla-temporary-toffoli}
\end{figure}

Many circuits involve computing and later uncomputing a Toffoli (e.g. addition).
Previously, it was believed that each Toffoli had a T-cost of 4 so the pair would have a T-cost of 8.
Our $|A\rangle$-investing construction state reduces the T-cost of the pair from 8 to 6.

@@

Third, taking another hint from \cite{Jones2013}, we can uncompute the ancilla by using a measure-and-correct process instead of with a mirror of the circuit that computed the ancilla.
To explain the uncomputatoin process, we start with a circuit that obviously performs the uncomputation and then perform simple circuit moves until we arrive at the intended destination.
Our starting point for the uncomputation circuit is simply a Toffoli gate, which obviously clears the ancilla since it was computed with a Toffoli and Toffolis are their own inverse.
Then, since the cleared ancilla is going to be discarded anyways, we are free to apply a Hadamard gate and a measurement to it after the Toffoli but before discarding it.
We then hop the Hadamard over the Toffoli, transforming it into a CCZ operation.
Next we re-arrange the CCZ so that the ancilla is one of the controls, which is possible because the controls and targets of a CCZ are interchangeable.
Finally, we invoke the deferred measurement principle \cite{Nielsen2009} to hop the Measurement over the CCZ, turning the quantum control into a classical control.
Ultimately, the Toffoli gate we started with has been turned into a CZ gate that we apply conditional on the result of measuring the ancilla.
The final circuit uses only Clifford operations, which have no T-cost.

With these three improvements applied, we are left with the key ingredients of our temporary AND gate.
We can initialize an ancilla storing the logical-AND of two qubits.
The creation process has a T-cost of 4.
Then we can keep the ancilla around, using it as a control for operations that would otherwise have been conditioned on both input qubits.
Finally, once the ancilla is no longer needed, we can erase it at no T-cost.

In diagrams, we draw the computation of the AND gate (i.e. the ancilla being initialized) as an ancilla wire emerging vertically from two controls then heading rightward (see \autoref{fig:compute-logical-and}).
We draw the uncomputation analogously, with the ancilla wire coming in from the left then merging vertically into the two controls that created it (see \autoref{fig:erase-logical-and}).
The initialized lifetime of the ancilla qubit corresponds to the wire coming out of and later merging into the control qubits.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &x &&\ctrl{1} &\qw & x  && &&          &&\targ    &\ctrl{2}&\qw     &\qTi &\qw     &\ctrl{2}&\targ     &\qw &\qw &\qw &\\
      &y &&\ctrl{1} &\qw & y  &&=&&          &&\targ    &\qw     &\ctrl{1}&\qTi &\ctrl{1}&\qw     &\targ     &\qw &\qw &\qw &\\
      &  &&         &\qw & xy && &&|A\rangle &&\ctrl{-2}&\targ   &\targ   &\qT  &\targ   &\targ   &\ctrl{-2} &\qH &\qS &\qw &\\
    }
  }
  \caption{
	Computing the logical-and of two qubits, with a T-cost of 4 and a T-depth of 1.
  }
  \label{fig:compute-logical-and}
\end{figure}

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &x  &&\ctrl{1} &\qw &x && &&\qw &\qw    &\ctrl{1} &\qw \\
      &y  &&\ctrl{1} &\qw &y &&=&&\qw &\qw    &\gate{Z} &\qw \\
      &xy &&\qw      &    &  && &&\qH &\meter &\cw \cwx \bullet &    \\
    }
  }
  \caption{
	Uncomputing the logical-and of two qubits, with a T-cost of 0.
  }
  \label{fig:erase-logical-and}
\end{figure}

We can perform a single Toffoli by computing the logical-and of two qubits, applying a CNOT from the logical-and onto the actual target, then uncomputing the logical-and.
This combined construction, shown in \autoref{fig:merge-use-erase-toffoli}, is equivalent to the one from \cite{Jones2013}.

>>>>

\title{Halving the cost of quantum addition}
\author{Craig Gidney}
\affiliation{Google, Santa Barbara, CA 93117, USA}
\email{craiggidney@google.com}

@@

\begin{abstract}
We improve the number of T gates needed to perform an $n$-bit adder from $8n + O(1)$ \cite{Cuccaro2004, Amy2013, AustinDiscussionsAndEmails2017} to $4n + O(1)$.
We do so via a ``temporary AND gate" construction, which uses four T gates to store the logical-AND of two qubits into an ancilla and zero T gates to later erase the ancilla.

Temporary AND gates are a generally useful tool when optimizing T-count.
They can be applied to integer arithmetic, modular arithmetic, rotation synthesis, the quantum Fourier transform, Shor's algorithm, Grover oracles, and many other circuits.
Because T gates dominate the cost of quantum computation based on the surface code, and the temporary AND gate is widely applicable, our constructions represent a significant reduction in projected costs of quantum computation.

In addition to our $n$-bit adder circuit with T-count of $4n + O(1)$, we show how to construct an $n$-bit controlled adder circuit with T-count of $8n + O(1)$, and a temporary adder that can be computed for the same cost as the normal adder but whose result can be kept until later uncomputed at no T-count.
\end{abstract}

@@

Our paper is divided into six sections.
In \autoref{sec:introduction}, we motivate the problem and introduce terminology.
\autoref{sec:review} discusses how previous work on optimizing T gate counts improved the cost of matched pairs of Toffolis from 14 to 8.
In \autoref{sec:invest}, we show how to improve the T-count from 8 to 6 by investing and later recovering an $|A\rangle$ state instead of consuming it to perform T gates.
\autoref{sec:temporary-and} expands on those ideas with an ancilla erasing trick that improves the T-count from 6 to 4.
This leads to \autoref{sec:circuit-constructions}, where we demonstrate how to use our constructions to create low T-count circuits for several addition tasks and discuss other possible applications of our constructions.
Finally, \autoref{sec:conclusion} summarizes our contributions, and proves that at least two T gates are needed to perform a temporary AND.

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ctrl{1}&\ctrl{1}&\ctrl{1}&\ctrl{1}&\qw && &&&\qw     &\qw     &\qw     &\ctrl{1}&\qw     &\qw     &\qw     &\qw &\\
      &\ctrl{1}&\ctrl{2}&\ctrl{3}&\ctrl{4}&\qw &&=&&&\qw     &\qw     &\qw     &\ctrl{1}&\qw     &\qw     &\qw     &\qw &\\
      &\targ   &\qw     &\qw     &\qw     &\qw && &&&\ctrl{3}&\ctrl{2}&\ctrl{1}&\targ   &\ctrl{1}&\ctrl{2}&\ctrl{3}&\qw &\\
      &\qw     &\targ   &\qw     &\qw     &\qw && &&&\qw     &\qw     &\targ   &\qw     &\targ   &\qw     &\qw     &\qw &\\
      &\qw     &\qw     &\targ   &\qw     &\qw && &&&\qw     &\targ   &\qw     &\qw     &\qw     &\targ   &\qw     &\qw &\\
      &\qw     &\qw     &\qw     &\targ   &\qw && &&&\targ   &\qw     &\qw     &\qw     &\qw     &\qw     &\targ   &\qw &\\
    }
  }
  \caption{
	The T-count of $N$ adjacent Toffolis sharing the same controls is $0 \cdot N + O(1)$.
	The marginal T-count is 0 because each additional Toffoli can be replaced with CNOTs framing a root Toffoli.
  }
  \label{fig:shared-controls}
\end{figure}

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ctrl{1} &\qw & &       & & &\ctrl{1}     &\ctrl{1} &\qw & &   & & &\qw &\qw &\qw      &\qw  &\ctrl{2} &\qw &\qw      &\qw  &\qw &\qw \\
      &\ctrl{1} &\qw & & \cong & & &\qS          &\ctrl{1} &\qw & & = & & &\qw &\qw &\ctrl{1} &\qw  &\qw      &\qw &\ctrl{1} &\qw  &\qw &\qw \\
      &\targ    &\qw & &       & & &\gate{Z}\qwx &\targ    &\qw & &   & & &\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\qw \\
    }
  }
  \caption{
	Starting with \autoref{fig:textbook-toffoli} then dropping T gates on the controls produces an operation with a T-count of 4 that performs the correct permutation.
	However, the operation introduces phase errors.
  }
  \label{fig:bad-phase-toffoli}
\end{figure}

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw && &&&\ctrl{1}     &\ctrl{1}  &\qw &\qw &\qw    &\qw &\qw     &\ctrl{1}&\ctrl{1}     &\qw & \\
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw &&=&&&\qS          &\ctrl{1}  &\qw &\qw &\qw    &\qw &\qw     &\ctrl{1}&\qSi         &\qw & \\
      &\targ   &\qw &    &\ldots &    &    &\targ   &\qw && &&&\gate{Z}\qwx &\targ     &\qw &    &\ldots &    &        &\targ   &\gate{Z}\qwx &\qw & \\
    }
  }
  \caption{
	When two Toffolis form a compute/uncompute pair, they can cancel each others' phase errors.
	Fixes the problem with the construction in \autoref{fig:bad-phase-toffoli}, and achieves a per-Toffoli T-count of 4 for paired Toffolis \cite{Barenco1995}.
  }
  \label{fig:cancelled-bad-phase-toffoli}
\end{figure}

A typical $n$-bit quantum adder will contain $2n + O(1)$ Toffoli gates, implying a naive T-count of $14n + O(1)$.
However, almost all of the Toffoli gates in the first half of an adder are uncomputed by Toffoli gates in the second half.
This allows the T gates on the controls of the Toffoli gates to be omitted, reducing their T-count from 7 to 4 and the T-count of addition to $8n + O(1)$.

Even if a Toffoli isn't paired with a second Toffoli that uncomputes its effects, it is still possible to perform the Toffoli with T-count of 4 by using an ancilla, a measurement, and a conditional fixup operation \cite{Jones2013} (see \autoref{fig:jones-toffoli}).

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ctrl{1} &\qw & &   & & &     &\qw &\qw &\qw      &\qw  &\ctrl{3} &\qw &\qw      &\qw  &\qw &\qw  &\qw       &\qw &\qw    &\ctrl{1}       &\qw \\
      &\ctrl{1} &\qw & & = & & &     &\qw &\qw &\ctrl{2} &\qw  &\qw      &\qw &\ctrl{2} &\qw  &\qw &\qw  &\qw       &\qw &\qw    &\gate{Z}       &\qw \\
      &\targ    &\qw & &   & & &     &\qw &\qw &\qw      &\qw  &\qw      &\qw &\qw      &\qw  &\qw &\qw  &\targ     &\qw &\qw    &\qw\cwx        &\qw \\
      &         &    & &   & & & \qO &\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\qSi &\ctrl{-1} &\qH &\meter &\cw\cwx\bullet & \\
    }
  }
  \caption{
	Toffoli construction with T-count of 4 from \cite{Jones2013}.
  }
  \label{fig:jones-toffoli}
\end{figure}

\section{Warm up: Reusing $|A\rangle$ states}
\label{sec:invest}

This paper started with an idea that improved the T-count of a single Toffoli gate from 7 to 6 in a surprising way, by performing the Toffoli indirectly instead of directly.
We started with a clean ancilla in the $|0\rangle$ state, applied a Toffoli to store the logical-AND of the two controls in the ancilla, used the ancilla to control a CNOT onto the intended target, then uncomputed the ancilla.

Initially, this indirect-Toffoli construction appears to have a T-count of 8.
However, as shown in \autoref{fig:indirect-toffoli}, the last T gate in the circuit is unnecessary.
In fact, it is actively harmful.
Dropping this T gate and the following Hadamard not only reduces the T-count from 8 to 7, it leaves the ancilla in an $|A\rangle$ state that can be consumed to perform a T gate elsewhere.
This improves the net T-count to 6.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ctrl{1} &\qw && &&&    &\qw &\qw &\qw      &\qw  &\ctrl{3} &\qw &\qw      &\qw  &\qw &\qw       &\qw &\qw &\qw      &\qw  &\ctrl{3} &\qw &\qw      &\qw  &\qw &\qw &&\\
      &\ctrl{1} &\qw &&=&&&    &\qw &\qw &\ctrl{2} &\qw  &\qw      &\qw &\ctrl{2} &\qw  &\qw &\qw       &\qw &\qw &\ctrl{2} &\qw  &\qw      &\qw &\ctrl{2} &\qw  &\qw &\qw &&\\
      &\targ    &\qw && &&&    &\qw &\qw &\qw      &\qw  &\qw      &\qw &\qw      &\qw  &\qw &\targ     &\qw &\qw &\qw      &\qw  &\qw      &\qw &\qw      &\qw  &\qw &\qw &&\\
      &         &    && &&&\qO &\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\ctrl{-1} &\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\qw 
          \gategroup{4}{6}{4}{10}{.7em}{--} \gategroup{4}{26}{4}{30}{.7em}{--} &|0\rangle & \\
      &         &    && &&&    &    |A\rangle &  & &     &         &    &         &     &    &          &    &    &         &    &         &     &         &     & |A\rangle
    }
  }
  \caption{
	Performing a Toffoli indirectly, by applying a pair of Toffolis to an ancilla and using the intermediate value to control a CNOT onto the actual target, appears to have a T-count of 8.
	However, the initial T gate is being used to compute an $|A\rangle$ state and the final T gate is being used to uncompute the $|A\rangle$ state.
    Instead of spending $|A\rangle$ states to perform T gates to make and unmake $|A\rangle$ states, we pass in and later recover a single $|A\rangle$ state.
	This ``investment" reduces the net T-count to 6.
  }
  \label{fig:indirect-toffoli}
\end{figure}

A Toffoli with a T-count of 6 isn't optimal, but the reduction from 7 was done in an unexpected way.
While searching for other circuits where this optimization might apply, we realized it would be useful when an initial Toffoli was later uncomputed by a second Toffoli.
Instead of computing and uncomputing the ancilla for the first Toffoli, then recomputing and re-uncomputing the ancilla for the second Toffoli, we simply kept the ancilla around until it was time to uncompute the Toffoli.
This halved the T-count of the pair from 12 to 6.
See \autoref{fig:ancilla-temporary-toffoli}.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw && &&&    &\qw     &\qw  &\ctrl{3}&\qw &\qw     &\qw  &\qw &\qw      &\qw &\qw &\qw    &\qw &\qw &\qw      &\qw &\qw &\qw     &\qw  &\ctrl{3}&\qw &\qw     &\qw &&\\
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw &&=&&&    &\ctrl{2}&\qw  &\qw     &\qw &\ctrl{2}&\qw  &\qw &\qw      &\qw &\qw &\qw    &\qw &\qw &\qw      &\qw &\qw &\ctrl{2}&\qw  &\qw     &\qw &\ctrl{2}&\qw &&\\
      &\targ   &\qw &    &\ldots &    &    &\targ   &\qw && &&&    &\qw     &\qw  &\qw     &\qw &\qw     &\qw  &\qw &\targ    &\qw &    &\ldots &    &    &\targ    &\qw &\qw &\qw     &\qw  &\qw     &\qw &\qw     &\qw &&\\
      &        &    &    &       &    &    &        &    && &&&\qA &\targ   &\qTi &\targ   &\qT &\targ   &\qTi &\qH &\ctrl{-1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{-1}&\qH &\qT &\targ   &\qTi &\targ   &\qT &\targ   &\qw &|A\rangle &\\
    }
  }
  \caption{
	Computing and uncomputing a Toffoli gate with a net T-count of 6.
	Invests an $|A\rangle$ state when computing the Toffoli, holds an ancilla qubit storing the logical-AND of the controls until it is time to uncompute the Toffoli, then recovers the $|A\rangle$ state.
  }
  \label{fig:ancilla-temporary-toffoli}
\end{figure}

Many circuits involve computing and later uncomputing a Toffoli (e.g. addition).
Previously, it was believed that each Toffoli had a T-count of 4 so the pair would have a T-count of 8.
Our $|A\rangle$-investing construction state reduces the T-count of the pair from 8 to 6.

@@

Third, taking another hint from \cite{Jones2013}, we can uncompute the ancilla by using a measure-and-correct process instead of with a mirror of the circuit that computed the ancilla.
To explain the uncomputatoin process, we start with a circuit that obviously performs the uncomputation and then perform simple circuit moves until we arrive at the intended destination.
Our starting point for the uncomputation circuit is simply a Toffoli gate, which obviously clears the ancilla since it was computed with a Toffoli and Toffolis are their own inverse.
Then, since the cleared ancilla is going to be discarded anyways, we are free to apply a Hadamard gate and a measurement to it after the Toffoli but before discarding it.
We then hop the Hadamard over the Toffoli, transforming it into a CCZ operation.
Next we re-arrange the CCZ so that the ancilla is one of the controls, which is possible because the controls and targets of a CCZ are interchangeable.
Finally, we invoke the deferred measurement principle \cite{Nielsen2009} to hop the Measurement over the CCZ, turning the quantum control into a classical control.
Ultimately, the Toffoli gate we started with has been turned into a CZ gate that we apply conditional on the result of measuring the ancilla.
The final circuit uses only Clifford operations, which have no T-count.

With these three improvements applied, we are left with the key ingredients of our temporary AND gate.
We can initialize an ancilla storing the logical-AND of two qubits.
The creation process has a T-count of 4.
Then we can keep the ancilla around, using it as a control for operations that would otherwise have been conditioned on both input qubits.
Finally, once the ancilla is no longer needed, we can erase it at no T-count.

In diagrams, we draw the computation of the AND gate (i.e. the ancilla being initialized) as an ancilla wire emerging vertically from two controls then heading rightward (see \autoref{fig:compute-logical-AND}).
We draw the uncomputation analogously, with the ancilla wire coming in from the left then merging vertically into the two controls that created it (see \autoref{fig:erase-logical-AND}).
The initialized lifetime of the ancilla qubit corresponds to the wire coming out of and later merging into the control qubits.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &x &&\ctrl{1} &\qw & x  && &&          &&\targ    &\ctrl{2}&\qw     &\qTi &\qw     &\ctrl{2}&\targ     &\qw &\qw &\qw &\\
      &y &&\ctrl{1} &\qw & y  &&=&&          &&\targ    &\qw     &\ctrl{1}&\qTi &\ctrl{1}&\qw     &\targ     &\qw &\qw &\qw &\\
      &  &&         &\qw & xy && &&|A\rangle &&\ctrl{-2}&\targ   &\targ   &\qT  &\targ   &\targ   &\ctrl{-2} &\qH &\qS &\qw &\\
    }
  }
  \caption{
	Computing the logical-AND of two qubits, with a T-count of 4 and a T-depth of 1.
  }
  \label{fig:compute-logical-AND}
\end{figure}

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &x  &&\ctrl{1} &\qw &x && &&\qw &\qw    &\ctrl{1} &\qw \\
      &y  &&\ctrl{1} &\qw &y &&=&&\qw &\qw    &\gate{Z} &\qw \\
      &xy &&\qw      &    &  && &&\qH &\meter &\cw \cwx \bullet &    \\
    }
  }
  \caption{
	Uncomputing the logical-AND of two qubits, with a T-count of 0.
  }
  \label{fig:erase-logical-AND}
\end{figure}

We can perform a single Toffoli by computing the logical-AND of two qubits, applying a CNOT from the logical-AND onto the actual target, then uncomputing the logical-AND.
This combined construction, shown in \autoref{fig:merge-use-erase-toffoli}, is equivalent to the one from \cite{Jones2013}.
